Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013

### 2017-07-04 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd

Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.

Design: A retrospective, repeated cross-sectional study of all deaths.

Setting: 39 US States and the District of Columbia.

Participants: Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

Primary outcome measure: We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.

Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Conclusions: Life expectancy for blacks and whites has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

@@

Need to remove one of these:
- Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.

@@

```{r print-plot-5, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

@@

```{r print-plot-6, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)

@@

`All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
```

>>>>

**Objective:** To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.

**Design:** A retrospective, repeated cross-sectional study of all deaths.

**Setting:** 39 US States and the District of Columbia.

**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

**Primary outcome measure:** We estimated trends in the gender-specific contribution of six major causes of death to the gap in life expectancy between black and whites and their 95% credible bands.

**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

**Conclusions:** Life expectancy for blacks and whites has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

@@

Need to remove one of these:

- Due to data suppression when death counts were small, we did not consider more specific causes of death such as homicide and drug overdose.

@@

```{r print-plot-5, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           scale_y_continuous(breaks = seq(-0.5, 1.5, by = 0.5), labels = seq(-0.5, 1.5, by = 0.5))

@@

```{r print-plot-6, fig.width = 12, fig.height = 10, echo = F}
#ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
`Non-communicable.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           scale_y_continuous(breaks = seq(-1, 1.5, by = 0.5), labels = seq(-1, 1.5, by = 0.5))

@@

`All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
```
